Supporting Information : The importance of a complete statistical description of dependence between variables for ecology and related fields

### 2018-12-06 Paper.Rmd
mostly edits to Intro

All branches of ecology and related areas study relationships among environmental and biological 
variables. However, commonly used correlation- and regression-based approaches to studying 
such relationships are limited - it is well understood that common statistical
approaches may provide only a very partial description of a relationship. 
For instance, datasets showing markedly different 
relationships may nevertheless have the same correlation coefficient value:  
Fig. \ref{fig_cop_pedag1}A-B show two such datasets. The panel A 
(respectively, B) variables of Fig. \ref{fig_cop_pedag1} are principally 
related in the left (respectively, right) portions of their distributions, a difference that may
have ecological significance, as discussed below, but is not captured by standard
correlation. Panels C and D 
were obtained from A and B through monotonic transformation (Appendix X for details),
<!--***DAN: Dan to go back and write the appendix-->
which alters the Pearson, but not the Spearman correlation. Although correlation
is only one of the numerous metrics for 
studying relationships between variables, other commonly used metrics also 
provide partial descriptions only.

Well-developed statistical tools do exist (REFS), and are applied widely in other fields 
(REFS), that provide a complete description of the dependence between two or more variables. 
The goal of this paper is to illustrate the potential for broader application of these
tools in ecology and related fields. The tools are based on the statistical concept of the *copula*, which we here introduce conceptually.

  a) then do it. some things to do include...
  b) the rank-based approach for data (use bivariate data)
  c) the inverse cdf approach for the copula of a bivariate distribution
  d) statement that copulas are also multivariate, though we won't explore that here (or maybe leave that for Discussion?)
  e) a one- or two-sentence description of sklars theorem

\begin{figure}[!h]
\begin{center}
\includegraphics[width=3.5in]{./Results/PedagogFig1.pdf}
\caption{(A-F, I, J) Diverse bivariate datasets, many of which have the same Pearson (P)
or Spearman (S) correlation coefficients. Panels C and E (respectively, D and F) were 
obtained from A (respectively, B) via monotonic functions that transformed the standard-
normal marginal distributions of A and B to gamma (C, D) or uniform (E, F) marginals. 
Details in Appendix X.\label{fig_cop_pedag1}}
\end{center}
\end{figure}
<!--***DAN: Dan to go back and add an appendix with details as it says. Make sure
to mention the shape and scale params of the gamma marginals in C, D were 2-->

Identifying and quantifying dependencies is the core of modelling in many applications, frequently found in biology and elsewhere. For many years, measuring and modeling dependencies has focused on the
Pearson correlation because of its ease in use. But correlation can provide little information about the underlying dependence structure in the case of asymmetric dependence (i.e., varying strength of association between two variables depending on their values) because it is a measure of linear association between two variables. It may be better to use rank-based correlation measures such as Kendall’s tau and Spearman’s rho instead of Pearson correlation when the marginal distribution of the variables are highly non-Gaussian. The copula [@joe2014_dependence; @nelsen2007_copula], a rank-based dependence measurement between two variables, transforms non-uniformly distributed marginals in original variables to uniformly distributed marginals when the joint distribution of ranked data is plotted. For example, in the first panel of Fig. \ref{fig_cop_pedag}, we showed the joint distribution of raw data (left) and copula data (right), where the marginals of raw data variables (x and y) were gamma distributed but in the copula plot those marginals had uniform distribution. The copula plot was obtained by scaling individual ranked variables by a factor $(n+1)$ where $n$ was the number of sampling points, so that the copula was confined within a unit box. From now, we will call this "normalized ranking". Using a "copula-based approach" as a dependence measure reflects more knowledge about the dependence structure than summarizing in just a single number, the linear correlation coefficient. It tells us about the tail-dependence of the variables, i.e., the association between extreme values of two variables. If smaller values of each variable are more correlated to each other, then the corresponding copula structure is said to have "lower tail-dependence" as shown in the first panel of Fig. \ref{fig_cop_pedag} copula plot (right). Similarly, an "upper tail-dependence" copula indicates stronger correlation between higher extreme variables in the raw data. Tail dependence occurs when the correlation between two variables increases as we get "further" in the tail (either or both) of the distribution. For example, in the middle panel of Fig. \ref{fig_cop_pedag}, different dependence structures, with same Kendall correlation $\tau = 0.5$, are shown which vary between dominantly upper tail dependent (lower tail - upper tail dependence being negative) to dominantly lower tail dependent (lower tail - upper tail dependence being positive) as we move along the row. Lower panel represents the similar scenario but with stronger correlation (Kendall's $\tau = 0.8$). So, the flexibility of copulae allows us to model more complex multivariate dependence structures by isolating marginal from joint distribution structures, with the capability of modeling a rich variety of tail behavior, such as tail dependence versus tail independence.<!--and reflection symmetry versus reflection asymmetry.--> To identify the kinds of copula structure found in real ecological data, we need to select candidate bivariate copulae from many existing parametric copula families. <!--A very flexible bivariate copula can itself become a welcome candidate in the pool of bivariate copulae to be selected from.--> Here, throughout our studies for different datasets we used a universal pool of 16 bivariate copula families (having exclusively lower, exclusively upper, both or neither tail dependence), described in the \nameref{Methods}.

Though a few examples have been found which shows copula approach to model dependencies in bioinformatics [@Kim2008], medical research [@Emura2016], a short period study of intersite dependence of rainfall data [@Serinaldi2008], a recent study on characterizing and assessing risks in drought management [@She2018] or monitoring changes in precipitation-extremes [@Goswami2018], the multivariate copula approach still, we believe, remains unexplored in ecology or related fields and thus has potential scope to better capture the evolution of ecological time series data.

@@

A model-selection approach based on likelihood and the Akaike Information Criterion (AIC) <!--and Bayesian Information Criterion (BIC)--> was used to help answer \hyperref[Intro]{\textbf{(Q1)}} from the \nameref{Intro}. For bivariate datasets 
$(x_i,y_i)$ for $i=1,...,n$ such as the soil C and N data, model selection involved several steps, 
most implemented using tools from the `VineCopula` package. <!--Code combining these tools is in the
[`OurBiCopSelect`](https://github.com/sghosh89/BIVAN/blob/master/OurBiCopSelect.R) function and its dependencies in the [`BIVAN`](https://github.com/sghosh89/BIVAN) repository. to be added in suppmat-->First, we produced ranks 
$(\tilde{u}_i,\tilde{v}_i)$ where $\tilde{u}_i$ is the rank of $x_i$ in the set 
$\{x_i:i=1,...,n\}$ and $\tilde{v}_i$ is the rank of $y_i$ in the set $\{y_i:i=1,...,n\}$ and 
we set $u_i=\tilde{u}_i/(n+1)$ and $v_i=\tilde{v}_i/(n+1)$; further work was with these "normalized ranked variables" $u_i$ 
and $v_i$. Ranks $\tilde{u}_i$ and $\tilde{v}_i$ were $1$, for the smallest element, up to $n$, for the largest. Second, we did a test of the independence of the $u_i$ and $v_i$. Subsequent model 
selection algorithms are ineffective if data could not be distinguished from independent data 
since most or all bivariate copula families considered include the independent copula in the 
family or as a boundary case. <!--The independence test was as implemented in the `BiCopIndTest`
function in `VineCopula`, the documentation of which reports the test uses the asymptotic normality of the statistic $|\tau|\sqrt{\frac{9n(n-1)}{2(2n+5)}}$, where $\tau$ is Kendall's tau statistic. to be added in suppmat-->If independence could be rejected ($0.05$ significance level), model selection proceeded. The third step of model selection was to fit multiple bivariate copula families to the ranked data $(u_i,v_i)$ via maximum likelihood, using the `BiCopEst` function in `VineCopula`. Families considered (with abbreviations here indicated) included the **N**ormal/Gaussian, **C**layton, **G**umbel, **F**rank, **J**oe, **BB1**, **BB6**, **BB7**, **BB8**, **S**urvival **C**layton , **S**urvival **G**umbel, **S**urvival **J**oe, **S**urvival **BB1**, **S**urvival **BB6**, **S**urvival **BB7** and **S**urvival **BB8** copulae. These copulae were chosen carefully so that they covered the whole range from exclusively lower (C, SG, SJ, SBB6, SBB8) or upper tail dependence (SC, G, J, BB6, BB8) to both tail dependence (BB1, SBB1, BB7, SBB7) in addition to two symmetric copulae (N, F). AIC and BIC values and accompanying model weights ($\text{AIC}_{\text{w}}$, $\text{BIC}_{\text{w}}$) were generated using standard formulas [@burnham2003_modelselection]. 
Descriptions of these copula families can be found in introductory texts on copulae [@nelsen2007_copula ; @joe2014_dependence]  and in `VineCopula` documentation. The lower- and upper-tail dependence of the best-fitting member of each family was also provided by `BiCopEst`. Finally, $\text{AIC}_{\text{w}}$ values for all fitted models were used to get
model-averaged lower- and upper-tail dependence values using standard model averaging 
formulas [@burnham2003_modelselection]; likewise for BIC.

@@

Left (or lower)- and right (or upper)-tail dependence quantify the extent to which two ranked
variables $u_i$, $v_i$ for $i=1,...,n$ 
(see \nameref{Model_selection} for defintions of $u_i$ and $v_i$)
are related to each other in the left and right tails, respectively, of their 
distributions [@nelsen2007_copula, @joe2014_dependence]. 
Our nonparametric study of tail dependence uses three
statistics (i.e., $\cor$, $\Ps$ and $\Dsq$) which can quantify the extent to which $u_i$ and $v_i$ are 
related in any part of their distributions. We here describe the 
statistics. The statistics were formulated with positively associated
variables in mind. All the variables we study are positively associated
when they are significantly associated. So definitions and associated 
diagrams will assume positive associations, though generalizations are 
possible. Given two bounds 
$0 \leq l_b < u_b \leq 1$, we define the parallel lines $u+v=2l_b$ and 
$u+v=2u_b$ which intersect the unit square in which the $u_i$ and $v_i$ 
fall (these are the blue and red lines, respectively, in 
Fig. \ref{fig_stat_image}). Our statistics quantify the dependence
between $u_i$ and $v_i$ in the region bounded by these lines.
Using $l_b=0$ and some value $u_b\leq 0.5$ can give information about 
dependence in the left parts of the distributions of $u$ and $v$, and 
using $u_b=1$ and $l_b \geq 0.5$ can give information about dependence in 
the right parts of the distributions.

@@

The difference $\cor_l-\cor_u$ is expected to be positive if 
dependence in the left halves of the distributions is stronger 
than dependence in their right halves; likewise for $\Ps_l-\Ps_u$.
This holds, for instance, for the Clayton copula [@nelsen2007_copula].
The difference $\Dsq_u-\Dsq_l$, in which the "upper" and
"lower" statistics occur in opposite order, is also expected to be 
positive under the same conditions. We use these three differences
to detect asymmetry of tail dependence. Likewise, for large datasets,
we consider $\cor_{l_b,u_b}-\cor_{1-u_b,1-l_b}$, 
$\Ps_{l_b,u_b}-\Ps_{1-u_b,1-l_b}$ and
$\Dsq_{1-u_b,1-l_b}-\Dsq_{l_b,u_b}$ with $l_b=0$ and $u_b$ close to $0$. We tested our statistics by testing whether $\cor_l-\cor_u$, $\Ps_l-\Ps_u$, and $\Dsq_u-\Dsq_l$ revealed the asymmetry of tail dependence known to be present in data simulated from various copulae (for details see Appendix \ref{SM-Test_npa_stat}).

@@

To test environmental effect on population copula from the model \ref{eq.AR1}, our approach is of two-fold : First, we calculated the parameter values of the noise-copula (for example, we chose Clayton with lower tail dependence, in this case the parameter of interest is $\theta$, which determines the strength of association of the variables [@nelsen2007_copula]) using `iRho` or `iTau` function from `copula` package for Spearman's Rho or Kendall's Tau, respectively. Then we simulated our model (\ref{eq.AR1}) and estimated parameters of the outcome bivariate population-copula, with last $500$ simulated points for each habitat patch (see Fig. \ref{fig_Cause4copula_params_GOF_500}). We also tested the goodness of fit of the estimated parameters for the population-copula and its standard error due to sample size. We reported p-values from goodness of fit tests for both statistics (Cramer-von Mises and Kolmogorov-Smirnov). p-values greater or equal to $0.05$ (above the dashed line in Fig. \ref{fig_Cause4copula_params_GOF_500}) were considered as a good fit of estimated parameters. Fig. \ref{SM-fig_Cause4copula_params_GOF_2500} represents similar feature as of Fig. \ref{fig_Cause4copula_params_GOF_500} but here we considered last $2500$ simulated population data points for our parameters-estimation, thus making our comparison more sensitive on basis of p-value threshold at $0.05$. (For code see [`Plotter_Cause4copula_GOF`](https://github.com/sghosh89/BIVAN/blob/master/Cause4copula.R) function).

>>>>

Very generally, all branches of ecology and related areas study relationships among environmental and biological 
variables. However, commonly used correlation- and regression-based 
approaches to studying 
such relationships are limited - it is well understood that common statistical
approaches provide only a very partial description of a relationship between variables. 
For instance, datasets showing markedly different 
relationships may nevertheless have the same correlation coefficient value:  
Fig. \ref{fig_cop_pedag1}A-B shows two such datasets. The variables of panel A 
(respectively, B) are principally 
related in the left (respectively, right) portions of their distributions, an asymmetric pattern of dependence in the distribution tails that may
have ecological significance, as discussed below, but is not captured by 
correlation. Panels C and D 
were obtained from A and B through monotonic transformation (Appendix X for details),
<!--***DAN: Dan to go back and write the appendix-->
which alters the Pearson, but not the Spearman correlation. Although correlation
is only one of the numerous metrics for 
studying relationships between variables, other commonly used metrics also 
provide partial descriptions only.

Well-developed statistical tools do exist, and are 
applied widely in other fields, that provide a demonstrably complete 
description of the dependence 
between two or more variables [@joe2014_dependence; @nelsen2006_copula]. 
The goal of this paper is to illustrate the potential for 
broader application of these
tools in ecology and related fields. The tools are based on 
the statistical concept of the *copula*, which we here 
introduce. From a data perspective, given 
a sample $(x_i,y_i)$, $i=1,2,\ldots,n$ (e.g., 
Fig. \ref{fig_cop_pedag1}, panel A or B), the associated copula
structure is revealed by the bivariate plot of $u_i$ versus $v_i$, where
$u_i$ is the rank of $x_i$ in the set of $x_j$ ($j=1,\ldots,n$), divided by
$n+1$; and $v_i$ is defined in the same way but using the $y_i$. See, 
for instance,
Fig. \ref{fig_cop_pedag1}E-F, which show the copula structure of 
A-B, respectively. Mathematically, if $(X,Y)$ is a continuous bivariate random
variable and the cummulative distribution functions (cdfs) of $X$ and $Y$ are 
$F_X(x)=P[X\leq x]$ and $F_Y(y)=P[Y\leq y]$, then the copula structure
of $(X,Y)$ is revealed by the random variable $(F_X(X),F_Y(Y))$.
This random variable takes values in the unit square, 
$[0,1] \times [0,1]$, and has uniform marginal distributions. Densities
corresponding to the copulas associated 
with Fig. \ref{fig_cop_pedag1}E-F are in Fig.
\ref{fig_cop_pedag1}G-H, respectively. Strictly speaking, 
the copula is actually defined as the bivariate 
cdf of $(F_X(X),F_Y(Y))$ rather than as this random variable itself, 
but this makes little conceptual difference. See @nelsen2006_copula and @joe2014_dependence
for mathematical details. An important theorem of Abe Sklar [@sklar1959_theorem] 
provides the result that copulas give a unique and complete 
representation of dependence
between variable: the original joint distribution of $(X,Y)$ 
can be reconstructed from the 
copula and the marginal distributions. Copula theory is also well developed for
multivariate distributions.

One way we may expect a copula approach  
to better reflect ecological relationships has to do with Liebig's law of the
minimum. If, for instance, the growth of a plant depends on soil 
nitrogen, N, as well
as on other factors, data on soil nitrogen and resulting growth rates may be 
expected to look more like Fig. \ref{fig_cop_pedag1}A or E than like 
Fig. \ref{fig_cop_pedag1}B or F. Nitrogen 
will control plant growth, producing a strong
relationship, only when it is limiting, in the left portion of the N and 
plant-growth-rate distributions. 
Nitrogen will be less related or unrelated to growth rates in the right tails of 
the distributions, as pictured in Fig. \ref{fig_cop_pedag1}A and E.

A second reason we may expect a copula-based approach to benefit 
ecological research is that such an approach may help illuminate 
causal mechanisms by which variables are related. For instance, if two species
X and Y are strong competitors, we may expect abundances of the two species
across quadrats or through time to be negatively related, as in 
Fig. \ref{fig_cop_pedag1}I-J. If X is plotted on the horizontal axis
and Y is plotted on the vertical, then the plot in Fig. \ref{fig_cop_pedag1}I 
suggests X is the dominant 
competitor: when X is abundant, Y is necessarily rare because it is suppressed 
by X; whereas when X is rare, Y
may be abundant or may also be relatively rare due to 
factors other than X. Alternatively, Fig. \ref{fig_cop_pedag1}J suggests that Y
is the dominant competitor.

A third reason we may expect a copula approach to benefit ecological
research is that prior work has demonstrated important copula 
stucture in the spatial dependence of environental 
variables [@Serinaldi2008; @She2018; @Goswami2018;
@li2013]. Such dependence tends to beget dependence across space
in fluctuations of populations or other ecosystem variables 
(this is called \emph{synchrony}) which are influenced by the environmental 
variables. This is called the \emph{Moran effect}. 
If copula structure is transmitted, 
possibly in modified form, 
from environmental to ecological variables, then we may expect important 
copula structure to be a common feature of dependencies through time 
between ecological variables measured in different locations. 
One may expect copula structure in environmental variables, in the form of 
accentuated tail-dependence, if
intense meteorological events are also spatially widespread.
We will explore this as well as other potential mechanisms for
copula structure in ecological variables.

A final \emph{a priori} reason we may expect copula structure to be important
in ecology has to do with spatially aggregated or averaged quantities. 
Many ecological variables of applied importance 
depend on the spatial average or sum of local 
quantities. For instance, regional methane of CO$_2$ 
fluxes are the sum of local fluxes; 
and the total economic value of a fishery is the sum of catches from 
local areas. We will explore how details, captured by the copula, of the 
nature of the dependence between local quantities can influence the spatial 
mean or sum and how this may influence human concerns and 
higher organizational levels in 
ecology. Illustrating the idea here, prior work demonstrated that the overall 
reliability of wind-generated electricity depends sensitively on details
of the dependence between wind speeds in multiple wind sites [@li2013]. 
Spatially aggregated ecological variables may show a similar dependence, with
potentially substantial consequences. For instance, if populations of a 
pest species in different locations are all mutually strongly
related to each other in their right tails, then local outbreaks tend strongly
to occur as part of regional epidemics. Similarly, if populations
of a crop resource species are strongly related in their left tails, then 
episodes of local scarcity tend to happen simultaneously, creating regional 
shortages.

<!--DAN: edit and add references to the following-->
Though a few examples have been found which shows copula approach to model dependencies in bioinformatics [@Kim2008], medical research [@Emura2016], a short period study of intersite dependence of rainfall data [@Serinaldi2008], a recent study on characterizing and assessing risks in drought management [@She2018] or monitoring changes in precipitation-extremes [@Goswami2018], the multivariate copula approach still, we believe, remains unexplored in ecology or related fields and thus has potential scope to better capture the evolution of ecological time series data.

\begin{figure}[!h]
\begin{center}
\includegraphics[width=3.5in]{./Results/PedagogFig1.pdf}
\caption{(A-F, I, J) Bivariate datasets showing diverse relationships between variables; many of these datasets nevertheless have the same Pearson (P)
or Spearman (S) correlation coefficients. Panels C and E (respectively, D and F) were 
obtained from A (respectively, B) via monotonic functions that transformed the standard-
normal marginal distributions of A and B to gamma (C, D) or 
uniform (E, F) marginals (details in Appendix X). (G, H) Contours 
of log-scale density functions of Clayton (G) and survival-Clayton 
(H) copulas with parameter 5 (see Methods). (I,J) Reflections
of E, F, respectively, about a vertical axis. Subscripts h and v 
stand for "horizontal" and "vertical". \label{fig_cop_pedag1}}
\end{center}
\end{figure}
<!--***DAN: Dan to go back and add an appendix with details as it says. Make sure
to mention the shape and scale params of the gamma marginals in C, D were 2-->

Using a "copula-based approach" as a dependence measure reflects more knowledge about the dependence structure than summarizing in just a single number, the linear correlation coefficient. It tells us about the tail-dependence of the variables, i.e., the association between extreme values of two variables. If smaller values of each variable are more correlated to each other, then the corresponding copula structure is said to have "lower tail-dependence" as shown in the first panel of Fig. \ref{fig_cop_pedag} copula plot (right). Similarly, an "upper tail-dependence" copula indicates stronger correlation between higher extreme variables in the raw data. Tail dependence occurs when the correlation between two variables increases as we get "further" in the tail (either or both) of the distribution. For example, in the middle panel of Fig. \ref{fig_cop_pedag}, different dependence structures, with same Kendall correlation $\tau = 0.5$, are shown which vary between dominantly upper tail dependent (lower tail - upper tail dependence being negative) to dominantly lower tail dependent (lower tail - upper tail dependence being positive) as we move along the row. Lower panel represents the similar scenario but with stronger correlation (Kendall's $\tau = 0.8$). So, the flexibility of copulae allows us to model more complex multivariate dependence structures by isolating marginal from joint distribution structures, with the capability of modeling a rich variety of tail behavior, such as tail dependence versus tail independence.<!--and reflection symmetry versus reflection asymmetry.--> To identify the kinds of copula structure found in real ecological data, we need to select candidate bivariate copulae from many existing parametric copula families. <!--A very flexible bivariate copula can itself become a welcome candidate in the pool of bivariate copulae to be selected from.--> Here, throughout our studies for different datasets we used a universal pool of 16 bivariate copula families (having exclusively lower, exclusively upper, both or neither tail dependence), described in the \nameref{Methods}.

@@

A model-selection approach based on likelihood and the Akaike Information Criterion (AIC) <!--and Bayesian Information Criterion (BIC)--> was used to help answer \hyperref[Intro]{\textbf{(Q1)}} from the \nameref{Intro}. For bivariate datasets 
$(x_i,y_i)$ for $i=1,...,n$ such as the soil C and N data, model selection involved several steps, 
most implemented using tools from the `VineCopula` package. <!--Code combining these tools is in the
[`OurBiCopSelect`](https://github.com/sghosh89/BIVAN/blob/master/OurBiCopSelect.R) function and its dependencies in the [`BIVAN`](https://github.com/sghosh89/BIVAN) repository. to be added in suppmat-->First, we produced ranks 
$(\tilde{u}_i,\tilde{v}_i)$ where $\tilde{u}_i$ is the rank of $x_i$ in the set 
$\{x_i:i=1,...,n\}$ and $\tilde{v}_i$ is the rank of $y_i$ in the set $\{y_i:i=1,...,n\}$ and 
we set $u_i=\tilde{u}_i/(n+1)$ and $v_i=\tilde{v}_i/(n+1)$; further work was with these "normalized ranked variables" $u_i$ 
and $v_i$. Ranks $\tilde{u}_i$ and $\tilde{v}_i$ were $1$, for the smallest element, up to $n$, for the largest. Second, we did a test of the independence of the $u_i$ and $v_i$. Subsequent model 
selection algorithms are ineffective if data could not be distinguished from independent data 
since most or all bivariate copula families considered include the independent copula in the 
family or as a boundary case. <!--The independence test was as implemented in the `BiCopIndTest`
function in `VineCopula`, the documentation of which reports the test uses the asymptotic normality of the statistic $|\tau|\sqrt{\frac{9n(n-1)}{2(2n+5)}}$, where $\tau$ is Kendall's tau statistic. to be added in suppmat-->If independence could be rejected ($0.05$ significance level), model selection proceeded. The third step of model selection was to fit multiple bivariate copula families to the ranked data $(u_i,v_i)$ via maximum likelihood, using the `BiCopEst` function in `VineCopula`. Families considered (with abbreviations here indicated) included the **N**ormal/Gaussian, **C**layton, **G**umbel, **F**rank, **J**oe, **BB1**, **BB6**, **BB7**, **BB8**, **S**urvival **C**layton , **S**urvival **G**umbel, **S**urvival **J**oe, **S**urvival **BB1**, **S**urvival **BB6**, **S**urvival **BB7** and **S**urvival **BB8** copulae. These copulae were chosen carefully so that they covered the whole range from exclusively lower (C, SG, SJ, SBB6, SBB8) or upper tail dependence (SC, G, J, BB6, BB8) to both tail dependence (BB1, SBB1, BB7, SBB7) in addition to two symmetric copulae (N, F). AIC and BIC values and accompanying model weights ($\text{AIC}_{\text{w}}$, $\text{BIC}_{\text{w}}$) were generated using standard formulas [@burnham2003_modelselection]. 
Descriptions of these copula families can be found in introductory texts on copulae [@nelsen2006_copula ; @joe2014_dependence]  and in `VineCopula` documentation. The lower- and upper-tail dependence of the best-fitting member of each family was also provided by `BiCopEst`. Finally, $\text{AIC}_{\text{w}}$ values for all fitted models were used to get
model-averaged lower- and upper-tail dependence values using standard model averaging 
formulas [@burnham2003_modelselection]; likewise for BIC.

@@

Left (or lower)- and right (or upper)-tail dependence quantify the extent to which two ranked
variables $u_i$, $v_i$ for $i=1,...,n$ 
(see \nameref{Model_selection} for defintions of $u_i$ and $v_i$)
are related to each other in the left and right tails, respectively, of their 
distributions [@nelsen2006_copula, @joe2014_dependence]. 
Our nonparametric study of tail dependence uses three
statistics (i.e., $\cor$, $\Ps$ and $\Dsq$) which can quantify the extent to which $u_i$ and $v_i$ are 
related in any part of their distributions. We here describe the 
statistics. The statistics were formulated with positively associated
variables in mind. All the variables we study are positively associated
when they are significantly associated. So definitions and associated 
diagrams will assume positive associations, though generalizations are 
possible. Given two bounds 
$0 \leq l_b < u_b \leq 1$, we define the parallel lines $u+v=2l_b$ and 
$u+v=2u_b$ which intersect the unit square in which the $u_i$ and $v_i$ 
fall (these are the blue and red lines, respectively, in 
Fig. \ref{fig_stat_image}). Our statistics quantify the dependence
between $u_i$ and $v_i$ in the region bounded by these lines.
Using $l_b=0$ and some value $u_b\leq 0.5$ can give information about 
dependence in the left parts of the distributions of $u$ and $v$, and 
using $u_b=1$ and $l_b \geq 0.5$ can give information about dependence in 
the right parts of the distributions.

@@

The difference $\cor_l-\cor_u$ is expected to be positive if 
dependence in the left halves of the distributions is stronger 
than dependence in their right halves; likewise for $\Ps_l-\Ps_u$.
This holds, for instance, for the Clayton copula [@nelsen2006_copula].
The difference $\Dsq_u-\Dsq_l$, in which the "upper" and
"lower" statistics occur in opposite order, is also expected to be 
positive under the same conditions. We use these three differences
to detect asymmetry of tail dependence. Likewise, for large datasets,
we consider $\cor_{l_b,u_b}-\cor_{1-u_b,1-l_b}$, 
$\Ps_{l_b,u_b}-\Ps_{1-u_b,1-l_b}$ and
$\Dsq_{1-u_b,1-l_b}-\Dsq_{l_b,u_b}$ with $l_b=0$ and $u_b$ close to $0$. We tested our statistics by testing whether $\cor_l-\cor_u$, $\Ps_l-\Ps_u$, and $\Dsq_u-\Dsq_l$ revealed the asymmetry of tail dependence known to be present in data simulated from various copulae (for details see Appendix \ref{SM-Test_npa_stat}).

@@

To test environmental effect on population copula from the model \ref{eq.AR1}, our approach is of two-fold : First, we calculated the parameter values of the noise-copula (for example, we chose Clayton with lower tail dependence, in this case the parameter of interest is $\theta$, which determines the strength of association of the variables [@nelsen2006_copula]) using `iRho` or `iTau` function from `copula` package for Spearman's Rho or Kendall's Tau, respectively. Then we simulated our model (\ref{eq.AR1}) and estimated parameters of the outcome bivariate population-copula, with last $500$ simulated points for each habitat patch (see Fig. \ref{fig_Cause4copula_params_GOF_500}). We also tested the goodness of fit of the estimated parameters for the population-copula and its standard error due to sample size. We reported p-values from goodness of fit tests for both statistics (Cramer-von Mises and Kolmogorov-Smirnov). p-values greater or equal to $0.05$ (above the dashed line in Fig. \ref{fig_Cause4copula_params_GOF_500}) were considered as a good fit of estimated parameters. Fig. \ref{SM-fig_Cause4copula_params_GOF_2500} represents similar feature as of Fig. \ref{fig_Cause4copula_params_GOF_500} but here we considered last $2500$ simulated population data points for our parameters-estimation, thus making our comparison more sensitive on basis of p-value threshold at $0.05$. (For code see [`Plotter_Cause4copula_GOF`](https://github.com/sghosh89/BIVAN/blob/master/Cause4copula.R) function).

### 2018-12-06 SupportingInformation.Rmd
mostly edits to Intro

Normal and Frank copula families 
are known to have symmetric tail dependence,
i.e., strengths of dependence in the lower and upper tails are the same, for all copulas
in these families [@nelsen2007_copula]. Clayton copulas are known to have stronger lower- than 
upper-tail dependence. And survival Clayton copulas, being rotations of Clayton copulas by 
$180^\circ$, have stronger upper- than lower tail dependence [@nelsen2007_copula].
Using the `iRho` function from the `copula` package, for each of the four families normal, 
Frank, Clayton, and survival Clayton, we obtained parameters for which the expected value of 
Spearman's $\rho$ was, separately, $0,0.1,0.2,...,0.9$. For each copula family and for each 
of the selected parameter values we generated  $n$ points, 100 times, and thereby computed 100 values of the 
statistics $\cor_l-\cor_u$, $\Ps_l-\Ps_u$, and $\Dsq_u-\Dsq_l$. This exercise 
was repeated, separately, 
for $n=1000$ and for $n=35$, these cases being chosen, respectively, to reflect asymptotic behavior of the 
statistics and their behavior on the small datasets which are common in some fields
of ecology. The mean and standard error of the three difference statistics were computed for 
each copula family, parameter value, and value of $n$, and results were plotted against 
Spearman's rho (Figures \ref{fig_stat_testing35} and \ref{fig_stat_testing1000}), 
the expectation being that the difference statistics would be positive for 
Clayton copulas and negative for survival Clayton copulas (except for the case of $\rho=0$, a 
boundary case of these families corresponding to independence), and not significantly different 
from $0$ (we used a $t$-test with a significance threshold of $0.05$) for Frank and normal 
copulas. These expectations were fullfilled (Figures \ref{fig_stat_testing35} and 
\ref{fig_stat_testing1000}).

>>>>

Normal and Frank copula families 
are known to have symmetric tail dependence,
i.e., strengths of dependence in the lower and upper tails are the same, for all copulas
in these families [@nelsen2006_copula]. Clayton copulas are known to have stronger lower- than 
upper-tail dependence. And survival Clayton copulas, being rotations of Clayton copulas by 
$180^\circ$, have stronger upper- than lower tail dependence [@nelsen2006_copula].
Using the `iRho` function from the `copula` package, for each of the four families normal, 
Frank, Clayton, and survival Clayton, we obtained parameters for which the expected value of 
Spearman's $\rho$ was, separately, $0,0.1,0.2,...,0.9$. For each copula family and for each 
of the selected parameter values we generated  $n$ points, 100 times, and thereby computed 100 values of the 
statistics $\cor_l-\cor_u$, $\Ps_l-\Ps_u$, and $\Dsq_u-\Dsq_l$. This exercise 
was repeated, separately, 
for $n=1000$ and for $n=35$, these cases being chosen, respectively, to reflect asymptotic behavior of the 
statistics and their behavior on the small datasets which are common in some fields
of ecology. The mean and standard error of the three difference statistics were computed for 
each copula family, parameter value, and value of $n$, and results were plotted against 
Spearman's rho (Figures \ref{fig_stat_testing35} and \ref{fig_stat_testing1000}), 
the expectation being that the difference statistics would be positive for 
Clayton copulas and negative for survival Clayton copulas (except for the case of $\rho=0$, a 
boundary case of these families corresponding to independence), and not significantly different 
from $0$ (we used a $t$-test with a significance threshold of $0.05$) for Frank and normal 
copulas. These expectations were fullfilled (Figures \ref{fig_stat_testing35} and 
\ref{fig_stat_testing1000}).
